---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# openFDA

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/openFDA)](https://CRAN.R-project.org/package=openFDA)
[![R-CMD-check](https://github.com/simpar1471/openFDA/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/simpar1471/openFDA/actions/workflows/R-CMD-check.yaml)
[![check-no-suggests](https://github.com/simpar1471/openFDA/actions/workflows/check-no-suggests.yaml/badge.svg)](https://github.com/simpar1471/openFDA/actions/workflows/check-no-suggests.yaml)
[![Codecov test coverage](https://codecov.io/gh/simpar1471/openFDA/graph/badge.svg)](https://app.codecov.io/gh/simpar1471/openFDA)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![pkgcheck](https://github.com/simpar1471/openFDA/workflows/pkgcheck/badge.svg)](https://github.com/simpar1471/openFDA/actions?query=workflow%3Apkgcheck)
<!-- badges: end -->

openFDA makes querying the [openFDA API](https://open.fda.gov/apis/) from R a breeze.
The API itself serves publicly available data from the FDA about foods, drugs, devices, and more.
This data includes data such as recall enforcement reports, adverse events, manufacturer details, and - again - even more!
Note that the data on openFDA has not been validated for clinical or production use.

## Installation

The easiest way to install openFDA is to get it from CRAN:
```r
install.packages("openFDA")
```

### Development version
```r
# install.packages("pak")
pak::pkg_install("simpar1471/openFDA")
```

## Using openFDA
```{r load_openFDA}
library(openFDA)
```

The full documentation for the API is online, so look at the [openFDA 
website](https://open.fda.gov/apis/) to get a full feel for the API itself.

Before using the package, you will need to generate one at the [openFDA webpage](https://open.fda.gov/apis/authentication/).
With an API key you can make 240 requests per minute, and are limited to 120,000 requests per day.
Without a key, you can only make 1000 requests a day.
For this package, if you try to run `openFDA()` without setting an API key, it will fail:

```{r store_api_key, echo = FALSE, eval = TRUE}
my_hidden_api_key <- get_api_key()
set_api_key(api_key = "")
```

```{r no_api_key, error = TRUE}
search <- openFDA(
  search = "openfda.generic_name:furosemide",
  limit = 5
)
```

That's not good.
To fix this error, get an API key from the webpage linked above.
Once you have your API key, use `set_api_key()` to store your API key.
Once you run this function, **openFDA** will 'know' your API key until the session ends.
Let's try using `openFDA()` again, after setting an API key:

```{r setting_api_key_fake, echo = TRUE, eval = FALSE}
# n.b. this API key is a fake, just for demonstration
set_api_key(api_key = "iHXeqlbXtlYEwX9MYLwyCBuiZxfn98Sj3oX2FNtSx")
```

```{r setting_api_key_real, echo = FALSE, eval = TRUE}
set_api_key(api_key = my_hidden_api_key)
```

```{r example1_query}
search <- openFDA(
  search = "openfda.generic_name:furosemide",
  limit = 5
)

search
```

The returned `httr2` response object contains JSON data from the API - you can use `httr2::resp_body_json()` to extract the JSON data as a nested list, then extract the fields you want.

```{r example2_json}
json <- httr2::resp_body_json(search)

json$results[[1]]$openfda$brand_name
json$results[[1]]$openfda$pharm_class_epc
```

I've found [purrr](https://purrr.tidyverse.org/) to be very useful for parsing this data quickly.

```{r example3_purrr}
purrr::map_chr(
  .x = json$results, 
  .f = \(result) purrr::pluck(result, "openfda", "manufacturer_name", 1)
)
```

## Other R packages for openFDA

The [openfda](https://github.com/rOpenHealth/openfda) package from `rOpenHealth` also wraps the openFDA API from R, and is available on GitHub.
It's got a pretty neat structure whereby you build up a query using individual functions for each parameter - though it's my personal preference to keep these as parameters to a single function.
It also makes results available in data frames, which is nice, but I think working with the response object and parsing the underlying JSON yourself permits more powerful interactions with the API.

There is also [FDAopenR](https://github.com/ck2136/FDAopenR/), which is all object-oriented.
I couldn't quite wrap my head around it, but it all appears to be in working order!
