---
title: "openFDA"
description: >
  Get up and running with accessing openFDA from R.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{openFDA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# save the built-in output hook
hook_output <- knitr::knit_hooks$get("output")

# set a new output hook to truncate text output
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x <- strsplit(x, split = "\n")[[1]]
    if (length(x) > n) {
      # truncate the output
      x <- c(head(x, n), "....\n")
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Using openFDA can be a breeze, if you know how to construct good queries.
This short guide will get you started with `{openFDA}`, and show you how to put together more complicated queries.
We start by loading openFDA.

```{r load_openFDA}
library(openFDA)
```

# The openFDA API
The openFDA API makes public FDA data available from a simple, public API.
Users of this API have access to FDA data on food, human and veterinary drugs, devices, and more.
You can read all about it at their [website](https://open.fda.gov/).
The API limits results sets to 1000 records.
To access over 1000 results, you'd usually need to manually submit multiple requests which skip past larger and larger chunks of the dataset ([see online for more info](https://open.fda.gov/apis/paging/)).

With this package, however, that can be handled for you.
By default, `openFDA()` will ask you if you want to retrieve all possible results.
However, asking for user input isn't possible when in a vignette.
Additionally, we don't want to do paging for these examples, as our searches get lots of results.
Therefore, to prevent paging, we set the package-level option `openFDA.paging` to `"never"`.
Now `openFDA()` will automatically **not** perform paging, even when it's possible.

```{r set-paging, echo = TRUE}
options(openFDA.paging = "never")
```

# A simple openFDA query
The simplest way to query the openFDA API is to identify the **endpoint** you want to use and provide other search terms.
For example, this snippet retrieves 2 records about **adverse events** in the **drugs** endpoint.
The empty search string (`""`) means the results will be non-specific.

```{r example1}
search <- openFDA(search = "",
                  endpoint = "drug-event",
                  limit = 2)
search
```

The warning given about pagination is because we searched on the whole dataset, but used `limit` to only pull out the first two results.
We can suppress this warning by setting `paging_verbosity` to `"quiet"`.
Now we won't page the whole dataset, because `paging == "never"`, and we won't get that warning either:

```{r example2}
search <- openFDA(search = "",
                  endpoint = "drug-event",
                  limit = 2,
                  paging = "never",
                  paging_verbosity = "quiet")
search
```

Instead of setting the `paging` and `paging_verbosity` arguments each time you
call the function, it is better to set the package-level option for these
behaviours. You can do this with any function that edits options:

```{r set-options}
options(
  openFDA.paging = "never",
  openFDA.paging_verbosity = "quiet"
)
```

## openFDA results
The function returns an `{httr2}` response object, with attached JSON data.
We use `httr2::resp_body_json()` to extract the underlying data.

```{r example_3a_json}
json <- httr2::resp_body_json(search)
```

If you don't specify a field to `count` on, the JSON data has two sections - `meta` and `results`.

### Meta
The `meta` section has important metadata on your results, which includes:

* `disclaimer` - An important disclaimer regarding the data provided by openFDA.
* `license` - A webpage with license terms that govern the openFDA API.
* `last_updated` - The last date at which this openFDA endpoint was updated.
* `results.skip` - How many results were skipped? Set by the `skip` parameter in `openFDA()`.
* `results.limit` - How many results were retrieved? Set by the `limit` parameter in `openFDA()`.
* `results.total` - How many results were there in total matching your `search` criteria?

```{r example_3b_json_meta}
json$meta
```

### Results

For non-`count` queries, this will be a set of records which were found in the endpoint and match your `search` term.

```{r example_3c_json_results, out.lines = 10}
json$results
```

### Results when `count`-ing
If you set the `count` query, then the openFDA API will not return full records.
Instead, it will count the number of records for each member in the openFDA field you specified for `count`.
For example, let's look at drug manufacturers in the [Drugs@FDA endpoint](https://open.fda.gov/apis/drug/drugsfda/) for `"paracetamol"`.

```{r example_4a_count_names, out.lines = 10}
count <- openFDA(search = "openfda.generic_name:ketamine",
                 endpoint = "drug-drugsfda",
                 count = "openfda.manufacturer_name.exact") |>
  httr2::resp_body_json()
count$results
```

### Using `count` to make a time series
You can count on fields with a date to create a time series.
This is demonstrated on the [openFDA
website](https://open.fda.gov/apis/timeseries/), where they count the number of received records per day in the `drug/event` endpoint.
With the openFDA package, you could do it in R like this:

```{r example_4b_count_timeseries, fig.alt = "A scatter plot of date (on x axis) against number of adverse events reported to the FDA (y-axis)."}
resp <- openFDA(endpoint = "drug-event", count = "receiptdate")
# Extract the counts from the response as a data frame
data <- httr2::resp_body_json(resp, simplifyDataFrame = TRUE) |>
  purrr::pluck("results")
data$date <- as.Date.character(data$time, format = "%Y%m%d")
# Sort on date column
data <- data[order(data$date), ]
# Restrict to data from 2000 to 2019
data <- data[data$date >= as.Date("2019-01-01") &
                     data$date < as.Date("2019-12-31"), ]
# Plot
plot(x = data$date, y = data$count,
     xlab = "Date", ylab = "Number of adverse events reported")
```

# Using search terms
We can increase the complexity of our query using the `search` parameter, which lets us search against specific openFDA API fields.
These fields are harmonised to different degrees in each API, which you will need to check [online](https://open.fda.gov/apis/openfda-fields/).

## Searching on one field
You can provide search strategies to `openFDA()` as single strings.
They are constructed as `[FIELD_NAME]:[STRING]`, where `FIELD_NAME` is the openFDA field you want to search on.
If your `STRING` contains spaces, you must surround it with double quotes, or openFDA will search against each word in the string.
So, for example, a search for drugs with the class `"thiazide diuretic"` should be formatted as `"openfda.pharm_class_epc:\"thiazide diuretic\""`, or the API will collect all drugs which have the words `"thiazide"` or `"diuretic"` in their established pharmacological class (EPC).

Note that there are helpers in the package (i.e. `construct_search_term()`) for assembling search terms; they are discussed below.

Let's do an unrefined search first:

```{r example_5a_single_search_term_unrefined}
search_unrefined <- openFDA(
  search = "openfda.pharm_class_epc:thiazide diuretic",
  endpoint = "drug-drugsfda",
  limit = 1
)
httr2::resp_body_json(search_unrefined)$meta$results$total
```

Let's compare this to our refined search, where we add double-quotes around the search term:

```{r example_5b_single_search_term_refined}
search_refined <- openFDA(
  search = "openfda.pharm_class_epc:\"thiazide diuretic\"",
  endpoint = "drug-drugsfda",
  limit = 1
)
httr2::resp_body_json(search_refined)$meta$results$total
```

As you can see, the unrefined search picked up `r httr2::resp_body_json(search_unrefined)$meta$results$total - httr2::resp_body_json(search_refined)$meta$results$total` more results, most of which would have probably been non-thiazide diuretics.

## Searching on multiple fields
The openFDA API also lets you search on various fields at once, to construct increasingly more specific queries.
You can either write your own complex search queries, or let the package do (some of) the heavier lifting for you.

### Write your own search term
Using the guides on the [openFDA website](https://open.fda.gov/apis/query-parameters/), you can put together your own query.
For example, the following query looks for up to 5 records which were submitted by Walmart and are taken orally.
We can use `{purrr}` functions to extract a brand name for each record.
Note that though a single record can have multiple brand names, we are choosing to only extract the first one.

```{r example_5c_supply_scalar_search_term}
search_term <- "openfda.manufacturer_name:Walmart+AND+openfda.route=oral"
search <- openFDA(search = search_term,
                  endpoint = "drug-drugsfda",
                  limit = 5)
json <- httr2::resp_body_json(search)
purrr::map(json$results, .f = \(x) {
  purrr::pluck(x, "openfda", "brand_name", 1)
})
```

### Let `openFDA()` construct the search term
In `openFDA()`, you can provide a named character vector with many field/search term pairs to the `search` parameter.
The function will automatically add double quotes (`""`) around your search terms, if you're providing field/value pairs like this.

```{r example_5d_format_scalar_search_term}
search <- openFDA(search = c("openfda.generic_name" = "amoxicillin"),
                  endpoint = "drug-drugsfda",
                  limit = 1)
paste("Number of results:", httr2::resp_body_json(search)$meta$results$total)
```

You can include as many fields as you like, as long as you only provide each field once.
By default, the terms are combined with an `OR` operator in `openFDA()`.
The below search strategy will therefore pick up all entries in [Drugs@FDA](https://open.fda.gov/apis/drug/drugsfda/) which are taken by mouth.

```{r example_5e_format_nonscalar_search_term}
search <- openFDA(search = c("openfda.generic_name" = "amoxicillin",
                             "openfda.route" = "oral"),
                  endpoint = "drug-drugsfda",
                  limit = 1)
paste("Number of results:", httr2::resp_body_json(search)$meta$results$total)
```

### Pre-construct a search term
You can use `format_search_term()` to pre-generate strings to pass to `openFDA()` as `search`:

```{r example_5f_nonscalar_search_term_OR}
format_search_term(
  c("openfda.generic_name" = "amoxicillin", "openfda.route" = "oral")
)
```

Sometimes you will want to combine multiple search terms with `AND` operators.
To do this, set the `mode` argument to `"and"`:
```{r example_5g_nonscalar_search_term_AND}
search_term <- format_search_term(
  c("openfda.generic_name" = "amoxicillin", "openfda.route" = "oral"),
  mode = "and"
)
search_term
```
```{r example_5h_nonscalar_search_term_run}
search <- openFDA(search = search_term,
                  endpoint = "drug-drugsfda",
                  limit = 1)
paste("Number of results:", httr2::resp_body_json(search)$meta$results$total)
```

## Wildcards
You can use the wildcard character `"*"` to match zero or more characters.
For example, we could take the prototypical ending to a common drug class - e.g. the **sartans**, which are angiotensin-II receptor blockers - and see which manufacturers are most represented in Drugs@FDA for this class.
When using wildcards, either pre-format the string yourself *without double-quotes* or use `format_search_term()` with `exact = FALSE`.
If you try to search with both double-quotes and the wildcard character, you will get a 404 error from openFDA.
In the code block below, we use a wildcard to find all drugs with generic names ending with `"sartan"`, and count on the manufacturer names.

```{r example_6_wildcards, out.lines = 15}
search_term <- format_search_term(c("openfda.generic_name" = "*sartan"),
                                  exact = FALSE)
search <- openFDA(search = search_term,
                  count = "openfda.manufacturer_name.exact",
                  endpoint = "drug-drugsfda",
                  paging = "always")
results <- httr2::resp_body_json(search, simplifyDataFrame = TRUE) |>
        purrr::pluck("results")
head(results, n = 10)
```

It looks like `"Alembic Pharmaceuticals"` is very active in this space - interesting!

# Large result sets
The openFDA API only returns up to 1000 results at one time.
To return a results set larger than 1000 records, a basic paging system is implemented.
On a per-call basis, this is controlled by the `paging` parameter in `openFDA()`.
To change this behaviour for the duration of an R session, edit the option `"openFDA.paging"`, as demonstrated above.

This argument must take one of these values:

* `"ask"` (default; only valid if R is running interactively) - ask before performing paging
* `"yes"` - always perform paging where necessary, but print a message to the console
* `"yes-quiet"` - always perform paging where necessary, but don't print a message to the console
* `"no"` - never perform paging, but print a message to the console
* `"no-quiet"` - never perform paging, but don't print a message to the console

When paging is performed, `openFDA()` returns a list of `{httr2}` response objects, instead of a single response:

```{r example_7_paging}
resps <- openFDA(search = "openfda.manufacturer_name:Pharma",
                 limit = 750,
                 endpoint = "drug-drugsfda",
                 paging = "always",
                 paging_verbosity = "verbose")
resps
```

# Other openFDA API features
This short guide does not cover all aspects of openFDA.
It is recommended that you go to the [openFDA API website](https://open.fda.gov/apis/) and check out the resources there to see information on:

* [Date ands ranges](https://open.fda.gov/apis/dates-and-ranges/)
* [Search for fields with missing values](https://open.fda.gov/apis/missing-values/)
